---
title: "Class notes and exercises"
output: html_document
date: "2025-02-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installed packages
```{r installed packages, results='hide'}
library(BiocManager)
library(IRanges)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)
library(purrr)
library(magrittr)
library(pheatmap)
library(textshape)
library(Rcpp)
library(matrixStats)
library(rtracklayer)
library(tidyverse)
library(DESeq2)
```


## Code and class exercises

### 2/19: Basics of reading and proccesing files in R
```{r 02/19}
# Load the data from tsv files
counts_matrix<- read.table("/scratch/Shares/rinnclass/MASTER_CLASS/STUDENTS/lupa9404/MASTER_CLASS/rna_seq_analysis/mouse_wt_long_timecourse_nf3.14/salmon.merged.gene_counts.tsv", header=TRUE, row.names=1)

# Create gene to symbol (g2s) dataframe that saves gene names and annotation from metadata in rownames
# Remember the format is: [rows, columns]. Metarows are pulled in by using rownames()
g2s<- data.frame(gene_id = rownames(counts_matrix), gene_symbol = counts_matrix[ , 1])

# Let's save this using write.csv
write.csv(g2s, "rna_seq_analysis/mouse_wt_long_timecourse_nf3.14/results/g2s.csv")

# Remove gene name column from matrix (matrices are numeric)
counts_matrix <- counts_matrix %>% select(-gene_name)

# Convert datframe into a matrix and round its values
counts_matrix_numerical <- as.matrix(counts_matrix) 
counts_matrix_rounded <- round(counts_matrix_numerical)

# Filter genes with no relevant counts (less than 1 for the sum of the row)
counts_filtered <- counts_matrix_rounded[rowSums(counts_matrix_rounded) > 1, ]

# Save progress in an .RData file
save(counts_filtered, counts_matrix, g2s, file = "rna_seq_analysis/mouse_wt_long_timecourse_nf3.14/results/count_files.RData")
```


### 2/21: Basics of R data proccesing + prep DESeq2 data
```{r 02/21}
# Load environment from previous class
load("rna_seq_analysis/mouse_wt_long_timecourse_nf3.14/results/count_files.RData")

# Make a column from the titles of the columns of the counts_matrix table
deseq_samples <- data.frame(
  sample_id = colnames(counts_matrix))

# Make a split_values variable with the split content of deseq_samples$sample_id 
split_values <- strsplit(deseq_samples$sample_id, "_")

# Make a list of the rime values and replicate values from the split_values variable
time_values <- sapply(split_values, function(x) x[[2]])
replicate_values <- sapply(split_values, function(x) x[[3]])

# Make columns in deseq_samples table with the time values and replicate values for each sample
deseq_samples$time_point <- time_values
deseq_samples$replicate <- replicate_values

# Factor columns with variables
deseq_samples$time_point <- factor(deseq_samples$time_point)
deseq_samples$replicate <- factor(deseq_samples$replicate)

# 
save(deseq_samples, counts_matrix, file = "rna_seq_analysis/mouse_wt_long_timecourse_nf3.14/results/deseq_samples.RData")
```

## Functions
- `read.table("<path>",header=TRUE, row.names=1)` : reads a table (such as tsv file) in R
- `data.frame(column1name = column1_values, column2name = column2_values)` : makes a dataframe with column names and values specified
- `write.csv(dataframe, "<path>/name.csv")` : writes a csv file from a dataframe
- `rownames(dataframe)` : returns the rows (meta-annotation) of a dataframe
- `select(variableFromDataframe)`: dplyr (from tidyverse) function to refer to a variable in a daframe (such as one of its column names)
- `as.matrix(dataframe)` : converts a dataframe into a matrix 
- `round(matrix)` : rounds all values in a matrix
- `rowSums(matrix)` : sums the values of a row in a matrix
- `save(object1, object2, file="<path>/filename.RData")` : saves files into an RData file type for later retrieval.
- `load(<path>/robject.RData)`: loads an .RData object and populates the enviroment with saved data
- `colnames(dataframe)`: retrieves the column names from a matrix-like object
- `strsplit(dataframe$vector_character, specification)`: Split the elements of a character vector into substrings according to the specification argument given (which is also a character).
- `sapply(dataframe, function)`: applies the function specified to the vector or dataframe passed in the argument.
- `factor(vector)`: encodes a vector as a factor

## General:
- **salmon.merged.gene_counts.tsv**: number of reads for each gene in the sample.

## To check
- How salmon works: https://combine-lab.github.io/salmon/getting_started/


