---
title: "installed_packages"
output: html_document
date: "2025-02-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Installed packages
```{r installed packages}
library(BiocManager)
library(IRanges)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)
library(purrr)
library(magrittr)
library(pheatmap)
library(textshape)
library(Rcpp)
library(matrixStats)
library(rtracklayer)
library(tidyverse)
library(DESeq2)
```


# Functions
- `read.table("<path>",header=TRUE, row.names=1)` : reads a table (such as tsv file) in R
- `data.frame(column1name = column1_values, column2name = column2_values)` : makes a dataframe with column names and values specified
- `write.csv(dataframe, "<path>/name.csv")` : writes a csv file from a dataframe
- `rownames(dataframe)` : returns the rows (meta-annotation) of a dataframe
- `select(variableFromDataframe)`: dplyr (from tidyverse) function to refer to a variable in a daframe (such as one of its column names)
- `as.matrix(dataframe)` : converts a dataframe into a matrix 
- `round(matrix)` : rounds all values in a matrix
- `rowSums(matrix)` : sums the values of a row in a matrix
- `save(object1, object2, file="<path>/filename.RData")` : saves files into an RData file type for later retrieval.

# General:
- **salmon.merged.gene_counts.tsv**: number of reads for each gene in the sample.

# To check
- How salmon works: https://combine-lab.github.io/salmon/getting_started/


